Towards a Verified Bzip2: A Functional Specification for Burrows-Wheeler

### 2019-05-03 draft_paper/draft_paper.tex
Finish permutation

There are multiple ways of defining \var{Permutation} as well: some of
them are more obviously correct, and some are more useful in certain
situations. The ``mathematical'' definition of when two lists $l$
and $l'$ are permutations of each other is:
\begin{itemize}[nosep]
  \item $\var{length}~l = \var{length}~l' = n$
  \item there exists a permutation function $p : \mathbb{N}_{<n} \to
    \mathbb{N}_{<n}$, where $p$ is a bijection, such that for all indices $i$,
    \begin{equation*}
      l_i = l'_{p(i)}
    \end{equation*}
\end{itemize}

In Coq, we can represent a finite function as a list of its outputs:
$f : \mathbb{N}_{<n} \to A$ is represented by the $\var{list}~A$
$[f(0),~f(1),\ldots,f(n-1)]$. This is consistent with the mathematical
notation where we denote by $2~1~0$ the permutation $p(0) = 2$, $p(1)
= 1$, $p(2) = 0$. So, $p$ here would be represented by $[2,~1,~0]$.

To restrict the range to $\mathbb{N}_{<n}$ we say that $\forall i,~i \in p
\implies ~i<n$, to make the function surjective, we strengthen that to
$\forall i,~i \in p \iff ~i<n$, and to make it injective we require that $p$ not
contain any duplicate entries. The inductive property \var{NoDup} from
the Coq standard library captures that requirement:
\begin{align*}
  \infer[\var{NoDup\_nil}]{\var{NoDup}~[]}{}
  &&
  \infer[\var{NoDup\_cons}]{\var{NoDup}~(x :: l)}{%
    x \notin l & \var{NoDup}~l
  }
\end{align*}
This leads to the following definition of the property
$\var{PermFun}~n~p$, which holds when $p : \var{list}~\mathbb{N}$
represents a permutation function on lists of length $n$:
\lstinputlisting[
  firstline=14, lastline=15
]{../theories/BWT/Sorting/PermFun.v}

@@

>>>>

The word \textit{permutation} has two related meanings that are often
used interchangably. We sometimes say that two lists are permutations of each
other if they contain all of the same elements, just appearing in a
different order. However, in mathematics we often say that a
permutation is actually the bijection on indices that defines the
correspondence between the two lists. For example, we say that $p$ is
the permutation that relates $l$ to $l'$ if for all indices $i$ and
$j$,
\begin{equation*}
  l_i = l'_{p(i)}
\end{equation*}
For consistency with Coq's built-in \var{Permuation} relation, in a
situation like this we will say that $l$ and $l'$ are permutations of
each other, and that $p$ is the \textit{permutation function} that
relates $l$ to $l'$. We say that two lists $l$ and $l'$ are
permutations of each other is precisely when
\vspace{-\parskip}
\begin{itemize}[nolistsep]
  \item $\var{length}~l = \var{length}~l' = n$
  \item there exists a permutation function $p : \mathbb{N}_{<n} \to
    \mathbb{N}_{<n}$, where $p$ is a bijection, such that for all indices $i$,
    \begin{equation*}
      l_i = l'_{p(i)}
    \end{equation*}
\end{itemize}

In Coq, we can represent a finite function as a list of its outputs:
$f : \mathbb{N}_{<n} \to A$ is represented by the $\var{list}~A$
$[f(0),~f(1),\ldots,f(n-1)]$. For permutations, this is consistent with the
notation where we denote by $2~1~0$ the permutation $p(0) = 2$, $p(1)
= 1$, $p(2) = 0$. So, $p$ here would be represented by $[2,~1,~0]$.

To restrict the range to $\mathbb{N}_{<n}$ we say that $\forall i,~i \in p
\implies ~i<n$, to make the function surjective, we strengthen that to
$\forall i,~i \in p \iff ~i<n$, and to make it injective we require that $p$ not
contain any duplicate entries. The inductive property \var{NoDup} from
the Coq standard library captures that requirement:
\begin{align*}
  \infer[\var{NoDup\_nil}]{\var{NoDup}~[]}{}
  &&
  \infer[\var{NoDup\_cons}]{\var{NoDup}~(x :: l)}{%
    x \notin l & \var{NoDup}~l
  }
\end{align*}

This leads to the following definition of the property
$\var{PermFun}~n~p$, which holds when $p : \var{list}~\mathbb{N}$
represents a permutation function on lists of length $n$:
\lstinputlisting[
  firstline=14, lastline=15,
]{../theories/BWT/Sorting/PermFun.v}
Then, we can define \var{apply}, which permutes a list according to a
given permutation:
\lstinputlisting[
  widthgobble=2, firstline=161, lastline=165,
]{../theories/BWT/Sorting/PermFun.v}
And finally, we can define \var{PermutationEx}, which holds between
two lists when they are permutations of each other.
\lstinputlisting[
  widthgobble=2, firstline=510, lastline=511,
]{../theories/BWT/Sorting/PermFun.v}

As with \var{Sorted}, we would also like an inductive definition of
\var{Permutation}. The Coq standard library includes one:
\begin{align*}
  \infer[\var{perm\_nil}]{\var{Permutation}~[]~[]}{}
  &&
  \infer[\var{perm\_skip}]{\var{Permutation}~(x :: l)~(x :: l')}{%
    \var{Permutation}~l~l'
  }
  \\
  \\
  \infer[\var{perm\_swap}]{\var{Permutation}~(x :: y :: l)~(y :: x :: l)}{}
  &&
  \infer[\var{perm\_trans}]{\var{Permutation}~l~l''}{
    \var{Permutation}~l~l' & \var{Permutation}~l'~l''
  }
\end{align*}

We will prove these two definitions equivalent, but first we confirm that our definition of \var{PermFun} ensures that .

We know we will
eventually need to reason inductively about permutation functions, so
we would like to define the relationship between applying a
permutation function $i :: p$ and applying $p$. Conceptually, the
first element of $\var{apply}~(i::p)~l$ will be $l_i$. The rest will
be the result of applying some new permutation $p'$, where everything
greater than $i$ has been shifted down by one, to some new list $l'$,
which has its $i$th element removed. So, we define \var{rem\_PermFun}
and \var{rem\_nth} respectively, hoping to prove that
\begin{equation*}
  \var{apply}~(i::p)~l =
  l_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~l)
\end{equation*}
\lstinputlisting[
  widthgobble=4, firstline=229, lastline=229,
]{../theories/BWT/Sorting/PermFun.v}
\vspace{\parskip}
\lstinputlisting[
  widthgobble=2, firstline=599, lastline=607,
]{../theories/BWT/Lib/List.v}

These two theorems define how \var{rem\_nth} affects indexing.
\begin{theorem}[nth\_lt\_rem\_nth, nth\_ge\_rem\_nth]
  For all lists $l$ and indices $i$ and $j$,
  \begin{equation*}
    j < i \implies (\var{rem\_nth}~i~l)_j = l_j
  \end{equation*}
  and
  \begin{equation*}
    j \ge i \implies (\var{rem\_nth}~i~l)_j = l_{j+1}
  \end{equation*}
\end{theorem}
\begin{proof}
  Both theorems follow by induction on $i$.
\end{proof}

As we expect, removing the $n$th element and sticking it on the front
preserves all elements.
\begin{theorem}[rem\_nth\_Perm]
  For all list $l$ and indices $i$
  \begin{equation*}
    \var{Permutation}~l~(l_i :: \var{rem\_nth}~i~l)
  \end{equation*}
\end{theorem}
\begin{proof}
  This follows from induction on $l$.
\end{proof}

\begin{theorem}[rem\_PermFun\_correct]
  For all permutation functions $p$, lists $l$, and indices $i$.
  \begin{equation*}
    \var{apply}~(i::p)~l =
    l_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~l)
  \end{equation*}
\end{theorem}
\begin{proof}
  Working backwards from our goal,
  \begin{align*}
    \var{apply}~(i::p)~l &=
    l_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~l) \\
    l_i :: \var{map}~(\lambda j.~l_j)~p &=
    l_i :: \var{map}~(\lambda j.~(\var{rem\_nth}~i~l)_j)~(\var{rem\_PermFun}~i~p) \\
    \var{map}~(\lambda j.~l_j)~p &=
    \var{map}~(\lambda j.~(\var{rem\_nth}~i~l)_j)~(\var{map}~(\lambda j. \bvar{if}~i < j~\bvar{then}~j-1~\bvar{else}~j)~p) \\
    \var{map}~(\lambda j.~l_j)~p &=
    \var{map}~(\lambda j.~(\var{rem\_nth}~i~l)_{\bvar{if}~i < j~\bvar{then}~j-1~\bvar{else}~j})~p \\
  \end{align*}
  From here, we need to prove that the mapping functions agree on all
  the elements of $p$, namely that for all $j \in p$
  \begin{equation*}
    l_j = (\var{rem\_nth}~i~l)_{\bvar{if}~i < j~\bvar{then}~j-1~\bvar{else}~j}
  \end{equation*}
  We know that $i \neq j$, because $j \in p$ and $\var{NoDup}~(i :: p)$.
  Therefore there are two cases, $i < j$ and $i > j$. When $i < j$, we have
  \begin{equation*}
    l_j = (\var{rem\_nth}~i~l)_{j-1}
  \end{equation*}
  by \var{nth\_ge\_rem\_nth}, and when $i > j$, we have
  \begin{equation*}
    l_j = (\var{rem\_nth}~i~l)_j
  \end{equation*}
  by \var{nth\_lt\_rem\_nth}.
\end{proof}

\begin{theorem}[apply\_correct]
  For all permutation functions $p$ and lists $l$,
  \begin{equation*}
    \var{Permutation}~(\var{apply}~p~l)~l
  \end{equation*}
\end{theorem}
\begin{proof}
  First we let $n = \var{length}~l$, so that we can reason by
  induction on $n$. The base case is uninteresting. In the inductive
  case, we have the induction hypothesis that for all lists of length
  $n$ and permutation functions on such lists,
  \begin{equation*}
    \var{Permutation}~(\var{apply}~p~l)~l
  \end{equation*}
  and we must show that
  \begin{equation*}
    \var{Permutation}~(\var{apply}~(i :: p)~(a :: l))~(a :: l)
  \end{equation*}

  We reason as follows, using the notation $\eqperm$ to
  chain together \var{Permutation}s:
  \begin{align*}
    \var{apply}~(i :: p')~(a :: l) &\eqperm
    (a :: l)_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~(a::l)) \\
    &\eqperm (a :: l)_i :: \var{rem\_nth}~i~(a::l) \\
    &\eqperm a :: l
  \end{align*}
  where the first step follows from \var{rem\_PermFun\_correct}, the
  second from the induction hypothesis, and the last from
  \var{rem\_nth\_Perm}.
\end{proof}

\begin{theorem}[PermutationEx\_iff]
  For all lists $l$ and $l'$,
  \begin{equation*}
    \var{Permutation}~l~l' \iff \var{PermutationEx}~l~l'
  \end{equation*}
\end{theorem}
\begin{proof}
  The $\impliedby$ direction follows from \var{apply\_correct}.

  ($\implies$) We reason by induction on the evidence of
  $\var{Permutation}~l~l'$. The empty case is simple.
  \begin{itemize}
  \item We have a permutation function $p$ from the induction
    hypothesis such that $\var{apply}~p~l = l'$, and we need to
    construct a permutation function $p'$ such that
    $\var{apply}~p'~(x::l) = x :: l'$. We let $p' = 0 ::
    \var{map}~(+1)~p$, shifting every index up by one and mapping
    the first element of the list to itself.
  \item We have to construct a permutation function $p$ such that
    $\var{apply}~p~(y :: x :: l) = x :: y :: l$. We let $p = 1 :: 0 ::
    \var{map}~(+2)~(\var{seq}~0~(\var{length}~l))$, leaving every
    element in place except for the first 2.
  \item We have permutation functions $p_1$ and $p_2$ from our two
    induction hypotheses, such that
    \begin{align*}
      \var{apply}~p_1~l = l' && \var{apply}~p_2~l' = l''
    \end{align*}
    We need to construct a $p'$ such that $\var{apply}~p'~l=l''$, so
    we let $p' = p_2 \comp p_1$. For the definition of composition of
    permutations, see \cref{appendix:perm_comp}. Then,
    \begin{equation*}
      \var{apply}~(p_2 \comp p_1)~l =
      \var{apply}~p_2~(\var{apply}~p_1~l) = l'' \qedhere
    \end{equation*}
  \end{itemize}
\end{proof}

For completeness, we will briefly mention a third definition of a
permutation relation that says that two lists are permutations of one
another when any element occurs an equal number of times in each:
\lstinputlisting[
  widthgobble=2, firstline=14, lastline=15,
]{../theories/BWT/Sorting/PermutationCount.v}
\var{count\_occ} counts the number of occurences of a given variable
using a decision procedure for Leibniz equality. Here that decision
procedure is \texttt{equiv\_dec} from an \var{EqDec} instance where we
require the equivalence relation itself by Leibniz equality.

\begin{theorem}[PermutationCount\_iff]
  For all lists $l$ and $l'$,
  \begin{equation*}
    \var{Permutation}~l~l' \iff \var{PermutationCount}~l~l'
  \end{equation*}
\end{theorem}
\begin{proof}
  The ($\implies$) direction follows from induction in the evidence of
  $\var{Permutation}~l~l'$.

  In the ($\impliedby$) direction, we reason by induction on $l$. In
  the inductive case, we have to show $\var{Permutation}~(a::l)~l'$.
  We know that $a \in l'$ because $\var{count\_occ}~a~l' =
  \var{count\_occ}~a~(a::l) > 0$. That means we can deconstruct $l'$
  as $L \dplus a :: R$, and
  \begin{equation*}
    L \dplus a :: R \eqperm a :: L \dplus R \eqperm a :: l
  \end{equation*}
  where the last step follows from the induction hypothesis.
\end{proof}

@@

\appendix

\section{Permutation Composition}
\label{appendix:perm_comp}
